```{r include=FALSE}
library(dplyr, quietly = TRUE)
library(tidyr, quietly = TRUE)
library(tidybayes, quietly = TRUE)
library(rethinking)
library(brms)
library(loo)
library(dagitty, quietly = TRUE)
library(ggdag, quietly = TRUE)
library(ggdist, quietly = TRUE)
library(patchwork, quietly = TRUE)
library(paletteer, quietly = TRUE)
```


# Conditional Manatees {#interactions}

## Building an interaction

Load the data, log transform the gdp measure, remove incomplete cases and create
a character column for Africa or Not Africa.

```{r}
data(rugged)
d <- rugged %>%
  filter(complete.cases(rgdppc_2000)) %>%
  mutate(log_gdp = log(rgdppc_2000),
         is_africa = if_else(cont_africa == 1, "Africa", "Not Africa"),
         is_africa = as.factor(is_africa))
dd <- d %>%
  drop_na(rgdppc_2000) %>%
  mutate(log_gdp_s = log_gdp / mean(log_gdp),
         rugged_s = scales::rescale(rugged),
         rugged_sc = scale(as.vector(rugged_s), center = TRUE, scale = FALSE))
# glimpse(dd)
```

and we use the following DAG, see overthinking box in introduction of section 8.1
for another possible DAG.

```{r}
dag <- ggdag::dagify(R ~ U, G ~ U + R, G ~ C) %>%
  ggdag::tidy_dagitty() %>%
  mutate(unobserved = name == "U")
# glimpse(dag)
ggplot(dag, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_point(aes(color = unobserved)) +
  geom_dag_text(color = "indianred") +
  geom_dag_edges() +
  scale_color_manual(values = c("TRUE" = "chartreuse2", "FALSE" = "khaki2")) +
  theme_dag() +
  theme(title = element_text(color = "midnightblue"),
        legend.position = "none") +
  labs(title = "rugged data dag")
```


```{r}
p1 <- dd %>%
  filter(grepl("^africa$", x = is_africa, ignore.case = TRUE)) %>%
  ggplot(aes(x = rugged_s, y = log_gdp_s)) +
  geom_smooth(method = "lm", formula = y ~ x, fill = "lightblue", color = "royalblue") +
  geom_point(color = "burlywood4") +
  theme_minimal() +
  labs(title = "African nations", x = "ruggedness (rescale)", 
       y = "log GDP (prop of mean)")
p2 <- dd %>%
  filter(!grepl("^africa$", x = is_africa, ignore.case = TRUE)) %>%
  ggplot(aes(x = rugged_s, y = log_gdp_s)) +
  geom_smooth(method = "lm", formula = y ~ x, fill = "burlywood1", color = "burlywood4") +
  geom_point(color = "royalblue") +
  theme_minimal() +
  labs(title = "Non-African nations", x = "ruggedness (rescale)",
       y = "log GDP (prop of mean)")
msg <- "Figure 8.2. Separate linear regressions inside and outside of Africa"
p1 + p2 + plot_annotation(title = msg)
```



### Making a rugged model

and split the data into countries from Africa and not.

```{r}
lst <- d %>%
  split(d$is_africa)
# str(lst)
```

and now creating a simple univariate model

$$
\log{(log\_gdp\_s_i)} \sim \mathcal{N}(\mu_i, \sigma) \\
\mu_i = \alpha + \beta \cdot rugged\_sc_i \\
\alpha \sim \mathcal{N}(1, 1) \\
\beta \sim \mathcal{N}(0, 1) \\
\sigma \sim \mathcal{Exp}(1)
$$

Now fit the model. Get the **rpior samples** by using `sample_prior = TRUE`

```{r}
a_file <- here::here("fits", "b08_01.rds")
# b8.1 <- readRDS(file = a_file)
b8.1 <- brm(
  data = dd,
  family = gaussian,
  log_gdp_s ~ 1 + rugged_sc,
  prior = c(
    prior(normal(1, 1), class = Intercept),
    prior(normal(0, 1), class = b),
    prior(exponential(1), class = sigma)
    ),
  sample_prior = TRUE,
  iter = 2000, warmup = 1000, chains = 4, cores = detectCores(),
  seed = 8
  )
b8.1 <- brms::add_criterion(b8.1, criterion = c("waic", "loo"))
saveRDS(b8.1, file = a_file)
posterior_summary(b8.1)
```







## Symmetry of interactions





## Continuous interactions




## Summary