```{r include=FALSE}
library(dplyr, quietly = TRUE)
library(tidyr, quietly = TRUE)
library(rethinking)
library(brms)
library(modelr)
library(simstudy)
```


# Appendix B - Gamma distribution {-}


```{r echo=FALSE}
message("TODO: I could not make the conversion from brms to rethinking work.,
        Even Kurtz's explanations, althoughbetter than McElreath, don't
        really solve this issue.")
```



This appendix sorts out the issues with gamma parameters.  They are covered by
@kurtz2020b but the different notations make the section a bit difficult to
follow.

Greek letters are great to represent parameters but, in this case, they will
be avoided to facilitate the discussion with the exception that $\mu = \text{mean}$
$\sigma=\text{standard deviation}$ because they are very well known conventions.

## Gamma distribution


The gamma distribution can be expressed in 3 ways

In terms of the $shape$, usually called $\alpha$ and $rate$, called $\beta$ 
the gamma distribution is

$$
\begin{align*}
\mathcal{Gamma}(y \mid\alpha, \beta) &= \frac{\beta^\alpha y^{\alpha-1} e^{-\beta y}}{\Gamma(\alpha)} \\
\mathcal{Gamma}(y \mid shape, rate) &= \frac{rate^{shape} y^{shape-1} e^{-rate \cdot y}}{\Gamma(shape)}
\end{align*}
$$

The $scale$ of the gamma distribution represents its dispersion. This is
mentioned in @elreath2020, section 12.1.2 top of p. 374.

$$
scale = dispersion = \frac{\sigma^2}{\mu^2}
$$
The $scale$ ad $rate$ are the reciprocal of each othe

$$
scale = \frac{1}{rate}
$$


As a result the gamma distribution can be expressed in terms of $shape$, called $\alpha$,
and $scale$, called $\theta$ as


$$
\begin{align*}
\mathcal{Gamma}(y \mid \alpha, \theta) &= \frac{y^{\alpha-1} e^{-\frac{y}{\theta}}}{\theta^\alpha\Gamma(\alpha)} \\
\mathcal{Gamma}(y \mid shape, scale) &= \frac{y^{shape-1} e^{-\frac{y}{scale}}}{scale^{shape}\Gamma(shape)}
\end{align*}
$$


Also, the gamma distribution can be expressed in terms of **mean** $\mu$ and
shape $\alpha$


$$
\begin{align*}
\mathcal{Gamma}(y \mid \mu, \alpha) &= \frac{(\frac{\alpha}{\mu})^\alpha}{\Gamma(\alpha)}
  y^{\alpha-1} \exp{(-\frac{\alpha y}{\mu})} \\
\mathcal{Gamma}(y \mid \mu, shape) &= \frac{(\frac{shape}\mu)^{shape}}{\Gamma(shape)}
  y^{shape-1} \exp{(-\frac{shape \cdot y}{\mu})}
\end{align*}
$$


## Usage

As explained in @elreath2020, section 12.1.2 p. 374, `rethinking` uses the 
`dgampois(x, mu, scale)` which uses the $mean = \mu$, named $\lambda$
and the $scale$, named $\phi$ in section 12.1.2.


On the other hand, as explained in @kurtz2020b, secion 12.1.2, `brms`
parameterizes the gamma likelihood in terms of $\mu$ and the $shape$ 
(called $\alpha$).

## Conversion

Now, `rethinking::ulam()` will give use results in terms of $\mu$ and $scale$ 
and `brms::brm()` will return results in terms of $\mu$ and $shape$. Therefore
the question is: *Given a $scale$ from `rethinking``, how do convert it to a
$shape$?*

My suggested answer is to use the `gammaGetShapeRate(mean, dispersion)` from the
`simstudy` package used extensively in this project to simulate priors.


```{r}
# rethinking::dampois()
```


## Example

The example is taken from the documentation for `rethinking::dgampois()`. 
It also happen to be used for practice 12H2 in @elreath2020.

```{r}
data("Hurricanes")
dataHurricanes <- Hurricanes
rm(Hurricanes)
```



```{r}
a_file <- here::here("fits", "mAppB1.rds")
mAppB1 <- readRDS(file = a_file)
# mAppB1 <- quap(
#     alist(
#         deaths ~ dgampois(mu , scale),
#         log(mu) <- a,
#         a ~ dnorm(0,100),
#         scale ~ dexp(1)
#     ),
#     data = dataHurricanes )
# saveRDS(mAppB1, file = a_file)
# mAppB1 <- quap(
#     alist(
#         deaths ~ dgampois(mu , scale),
#         log(mu) <- a + b*femininity,
#         a ~ dnorm(0,100),
#         b ~ dnorm(0,1),
#         scale ~ dexp(1)
#     ),
#     data = dataHurricanes )
# saveRDS(mAppB1, file = a_file)
```


```{r}
precis(mAppB1)
```

and the fit with `brm`

```{r}
a_file <- here::here("fits", "bAppB1.rds")
bAppB1 <- readRDS(file = a_file)
# bAppB1 <- brm(data = dataHurricanes,
#       family = negbinomial,
#       deaths ~ 1,
#       prior = c(prior(normal(0, 100), class = Intercept),
#                 prior(gamma(0.01, 0.01), class = shape)),  # alpha
#       cores = detectCores())
# bAppB1 <- brm(data = dataHurricanes,
#       family = negbinomial,
#       deaths ~ 1 + femininity,
#       prior = c(prior(normal(0, 100), class = Intercept),
#                 prior(normal(0, 1), class = b),
#                 prior(gamma(0.01, 0.01), class = shape)),  # alpha
#       cores = detectCores())
# saveRDS(object = bAppB1, file = a_file)
```

which gives these estimates

```{r}
posterior_summary(bAppB1)
```

The results form `quap` and `brm` are similar except the $scale$ and $shape$.

```{r}
rethink <- list()
rethink$mu <- exp(precis(mAppB1)["a", "mean"])
# rethink$mu <- exp(precis(mAppB1)["a", "mean"] + precis(mAppB1)["b", "mean"])
rethink$mu
rethink$scale <- precis(mAppB1)["scale", "mean"]
rethink$scale
```

Knowing from above that $scale = dispersion$ then we use
`gammaGetShapeRate(mean, dispersion)` to obtain the shape

```{r}
posterior_summary(bAppB1)["shape", "Estimate"]
brm <- list()
b1 <- gammaGetShapeRate(mean = rethink$mu, dispersion = rethink$scale)
b1
b2 <- gammaGetShapeRate(mean = rethink$mu, dispersion = 1 / rethink$scale)
b2
```

```{r}
brm <- list()
brm$mean <- exp(posterior_summary(bAppB1)["b_Intercept", "Estimate"])
brm$shape <- posterior_summary(bAppB1)["shape", "Estimate"]
gammaGetMeanDispersion <- function(shape, rate) {
  stopifnot(shape > 0, rate > 0)
  dispersion <- 1 / shape
  mean <- shape / rate
  list("mean" = mean, "dispersion" = dispersion)
}
t <- gammaGetMeanDispersion(shape = brm$shape, rate = brm$mean)
t
gammaGetShapeRate(mean = t$mean, dispersion = t$dispersion)
```

